\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap

### 2017-03-14 TeX/paper.tex
simulation of networks

When it comes to the interpretation of the coefficients one can show that
\frac{\mathbb{P}_{\theta}(Y_{ij}=1 | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}=0 | Y_{ij}^c=G_{ij}^c)}
= \exp(\theta_1 (\Delta_1 G)_{ij}) \cdot ... \cdot \exp( \theta_q (\Delta_q G)_{ij})
\label{Inter2}
where the condition $Y_{ij}^c=G_{ij}^c$ is short for: $Y_{pq}=g_{pq}$ for all $(p,q)\in \{1,\dots,N\}^2$ with $(p,q) \neq (i,j)$ and refers to the rest of the network.
The expression $(\Delta G)_{ij}:=\Gamma(G_{ij}^+) - \Gamma(G_{ij}^-)$ is called the \textit{change statistic}. $G_{ij}^+$ emerges from $G$, while assuming $g_{ij}=1$ and $G_{ij}^-$ emerges from $G$, while assuming $g_{ij}=0$. This means that the $k$th component of $(\Delta G)_{ij}$ captures the difference between the networks $G_{ij}^+$ and $G_{ij}^-$ on the $k$th integrated statistic in the model.\\[0.3cm]
Equation (\ref{Inter2}) now enables a \textit{ceteris-paribus analysis} of the parameters in the model:\\
If the $k$th change statistic $(\Delta_k G)_{ij}$ increases one unit to $(\Delta_k G)_{ij}+1$, while all the other change statistics remain unchanged, the odds of occurrence of edge $(i,j)$, conditional on the rest of the network, is multiplied by the factor $\exp(\theta_k)$. This means that for $\theta_k$, $k \in \{1, \ldots , q\}$ the conditional odds of occurence increase if $\theta_k > 0 $, decrease if $\theta_k < 0 $ and stay the same if $\theta_k = 0 $. 
Therefore, the interpretation of the parameter happens very similar to logistic regression analysis. 
Notice that for the sake of simplicity we did not condition on exogenous network statistics in this chapter.

@@

\noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals. Furthermore, this method also adopts the advantage of the MPLE of being consistent. However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit especially if the size of the examined network is extremely large.

>>>>

%When it comes to the interpretation of the coefficients one can show that
%\begin{equation}
%\frac{\mathbb{P}_{\theta}(Y_{ij}=1 | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}=0 %| Y_{ij}^c=G_{ij}^c)}
%= \exp(\theta_1 (\Delta_1 G)_{ij}) \cdot ... \cdot \exp( \theta_q (\Delta_q G)_{ij})
%\label{Inter2}
%\end{equation}
%where the condition $Y_{ij}^c=G_{ij}^c$ is short for: $Y_{pq}=g_{pq}$ for all $(p,q)\in \{1,\dots,N\}^2$ with $(p,q) \neq (i,j)$ and refers to the rest of the network.
%The expression $(\Delta G)_{ij}:=\Gamma(G_{ij}^+) - \Gamma(G_{ij}^-)$ is called the \textit{change statistic}. $G_{ij}^+$ emerges from $G$, while assuming $g_{ij}=1$ and $G_{ij}^-$ emerges from $G$, while assuming $g_{ij}=0$. This means that the $k$th component of $(\Delta G)_{ij}$ captures the difference between the networks $G_{ij}^+$ and $G_{ij}^-$ on the $k$th integrated statistic in the model.\\[0.3cm]
%Equation (\ref{Inter2}) now enables a \textit{ceteris-paribus analysis} of the parameters in the model:\\
%If the $k$th change statistic $(\Delta_k G)_{ij}$ increases one unit to $(\Delta_k G)_{ij}+1$, while all the other change statistics remain unchanged, the odds of occurrence of edge $(i,j)$, conditional on the rest of the network, is multiplied by the factor $\exp(\theta_k)$. This means that for $\theta_k$, $k \in \{1, \ldots , q\}$ the conditional odds of occurence increase if $\theta_k > 0 $, decrease if $\theta_k < 0 $ and stay the same if $\theta_k = 0 $. 
%Therefore, the interpretation of the parameter happens very similar to logistic regression analysis. 
%Notice that for the sake of simplicity we did not condition on exogenous network statistics in this chapter.

No matter which kind of MCMC algorithm is used, the basic idea is the following:
\\One constructs a \textit{Markov chain} $(Y_t)_{t \in \mathbb{N}}$ on the set of all possible networks $\mathcal{A}(N)$ of $N$ nodes, whose \textit{stationary distribution} is in conformity with the distribution $\mathbb{P}_{\theta_0}$. One can show that every single realization (or \textit{trajectory}) 
%$$(A_t)_{t \in \mathbb{N}}:= (Y_t(\omega))_{t \in \mathbb{N}}$$
of this stochastic process accomplishes the convergence result (\ref{konver}) (for this version of the \textit{Law of big numbers for Markov chains} we reference Meyn and Tweedie \cite{Meyn.2009}). As a result, sub-sequences of $(A_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
Snijders \cite{Snijders.2002b} provides a Metropolis Hastings algorithm on how to simulate networks:
Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
\begin{enumerate}
\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
\item Compute the value
$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
\begin{itemize}
\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
\item $Z=1$, define $G^{(k+1)}$ via
$$g_{pq}^{(k+1)}=\begin{cases}
1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
\end{cases}$$
\end{itemize}
\item Start at step 1 with $G^{(k+1)}$.
\end{enumerate}
The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$. Since the orignial matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.

@@

\noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals. Furthermore, this method also adopts the advantage of the MPLE of being consistent. However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit, especially if the size of the examined network is extremely large.
